Compositional Quantitative Resource Analysis of C Programs

### 2014-06-16 paper/main.tex
more work on the catalog

Example \emph{fig\_4\_2} from [GulwaniPLDI09] is quite involved.
Amortized reasoning helps to understand how we derive the bound $4 +
4|[0, n]|$.  We start with potential $6 + 4|[0, n]|$ and use the
constant potential to pay for the two first assignments.  We use the
remaining potential $4 + 4|[0, n]|$ and the fact that $vb=0$ to
establish the potential $4 + 4|[0, n]| + 2|[0,vb]|$ that serves as a
loop invariant.  In the \emph{if branch} of the conditional, we use
the constant potential $4$ of the invariant to pay for the two
assignments ($2$) and the potential of $|0,vb|$ ($2$).  Since we also
know that $|[0, n]|>0$ we obtain constant potential $4$ and establish
the loop invariant again.  In the \emph{else branch}, we use the
potential $2|[0,vb]|$ and the fact $vb>0$ to obtain $2$ potential
units to pay for the assignment.  Since we have $4$ constant potential
units left after the loop, we can subtract them from the initial
potential $6 + 4|[0, n]|$.

In example \emph{fig\_4\_4} it is essential that $m$ is positive.
That ensures that we can obtain constant potential for the interval
size $|[0,i]|$ in the else branch of the conditional since $|[0,i]|$
decreases.  Example \emph{fig\_4\_5} is one of the three examples that
we can not handle automatically.  The reasoning is based on the fact
that the boolean value of the test \code{dir==fwd} does not change
during the iteration of the loop.  As a result, the variable $i$ is
either counted down to $0$ or up to $n$.  Our tool cannot handle
example \emph{fig\_4\_5} because we don't do path sensitive reasoning.
Note however that it would be more efficient to move the test
\code{dir==fwd} outside of the loop (this would be also done by an
optimizing compiler).  The resulting program can be analyzer by our
tool.

Example \emph{ex1} from [GulwaniPLDI10] specifically focusses on the
code in \emph{if} statement.  So we use the \emph{tick metric} and
insert \code{_tick(1)} inside the if statement to derive a bound on
the number of times the code in the if statement is executed.  Note
that we cannot derive a bound for the whole program since the outer
loop is executed a quadratic number of times.  Nevertheless it is
straightforward to derive a bound on the number of ticks using the
amortized approach: In the if statement we know that $n>0$ and assign
$n=n-1$.  So we can use the potential of the interval size $|[0,n]|$
to pay for the tick.

We cannot analyze example \emph{ex2} for which previous work reported
a linear bound [GulwaniPLDI10].  The reason is that this example is
not terminating since the inner loop is not bounded.  The same is true
for example \emph{ex5} from the same paper, which is not listed here.
Example \emph{ex2} is similar to example \emph{nested\_single} and
\emph{ex3} is a variant of example \emph{t47}.

@@

Example \emph{t37} illustrates that the analyzer is able to perform
inter-procedural size tracking.  The function \code{copy} adds the
argument $x$ to the argument $y$ if $x$ is positive.  However, this
addition is done in steps of $1$ in each recursive call.  The function
\code{count\_down} recursively decrements its argument down to $0$.
The derived bound $1 + 4|[0, x]| + 2|[0, y]|$ is for the function
\code{_main} in which we first add $x$ to $y$ using the function
\code{copy} and then count down the variable $y$ using the function
\code{count\_down}.  The derived bound is tight.

Example \emph{t39} uses mutual recursion.  The function
\code{count\_down} is similar to the function with the same name in
example \emph{t37}.  However, we do not count down to $0$ but to a
variable $y$ that is passed as an argument and we call the function
\code{count\_up} afterwards.  The function \code{count\_up} is dual to
\code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
\code{count\_down}.  For the function $\_main$ that calls
\code{count\_down(y,z)}, the analyzer computes the tight bound $1.67 +
1.33 |[z,y]|$.

@@

$XXX$
\\[.7\baselineskip]
      {\bf GCD}
    \end{center}
  \end{minipage}

>>>>

Example \emph{fig\_4\_2} from [GulwaniPLDI09] is quite involved.
Amortized reasoning helps to understand how we derive the bound $4 +
4|[0, n]|$.  We start with potential $6 + 4|[0, n]|$ and use the
constant potential to pay for the two first assignments.  We use the
remaining potential $4 + 4|[0, n]|$ and the fact that $vb=0$ to
establish the potential $4 + 4|[0, n]| + 2|[0,vb]|$ that serves as a
loop invariant.  In the \emph{if branch} of the conditional, we use
the constant potential $4$ of the invariant to pay for the two
assignments ($2$ units) and the potential of $|[0,vb]|$ ($2$ units).
Since we also know that $|[0, n]|>0$ we obtain constant potential $4$
and establish the loop invariant again.  In the \emph{else} branch, we
use the potential $2|[0,vb]|$ and the fact $vb>0$ to obtain $2$
potential units to pay for the assignment.  Since we have $4$ constant
potential units left after the loop, we can subtract them from the
initial potential $6 + 4|[0, n]|$.

In example \emph{fig\_4\_4} it is essential that $m$ is positive.
That ensures that we can obtain constant potential for the interval
size $|[0,i]|$ in the \emph{else} branch of the conditional since
$|[0,i]|$ decreases.  Example \emph{fig\_4\_5} is one of the three
examples that we can not handle automatically.  The execution is
bounded because the boolean value of the test \code{dir==fwd} does not
change during the iteration of the loop.  As a result, the variable
$i$ is either counted down to $0$ or up to $n$.  Our tool cannot
handle example \emph{fig\_4\_5} because we don't do path sensitive
reasoning.  Note however that it would be more efficient to move the
test \code{dir==fwd} outside of the loop (this would be also done by
an optimizing compiler).  The resulting program could be analyzer by our
tool.

Example \emph{ex1} from [GulwaniPLDI10] specifically focusses on the
code in the \emph{if} statement.  So we use the \emph{tick metric} and
insert \code{_tick(1)} inside the if statement to derive a bound on
the number of times the code in the if statement is executed.  Note
that we cannot derive a bound for the whole program since the outer
loop is executed a quadratic number of times.  Nevertheless it is
straightforward to derive a bound on the number of ticks using the
amortized approach: In the if statement we know that $n>0$ and assign
\code{n=n-1}.  So we can use the potential of the interval size $|[0,n]|$
to pay for the tick.

We cannot analyze example \emph{ex2} for which previous work reported
a linear bound [GulwaniPLDI10].  The reason is that this example is
not terminating since the inner loop is not bounded.  The same is true
for example \emph{ex5} from the same paper, which is not listed here.
Finally, example \emph{ex2} is similar to example
\emph{nested\_single}, and \emph{ex3} is a variant of example
\emph{t47}.

@@

Example \emph{t37} illustrates that the analyzer is able to perform
inter-procedural size tracking.  The function \code{copy} adds the
argument $x$ to the argument $y$ if $x$ is positive.  However, this
addition is done in steps of $1$ in each recursive call.  The function
\code{count\_down} recursively decrements its argument down to $0$.
The derived bound $1 + 4|[0, x]| + 2|[0, y]|$ is for the function
\code{\_main} in which we first add $x$ to $y$ using the function
\code{copy} and then count down the variable $y$ using the function
\code{count\_down}.  The derived bound is tight.

Example \emph{t39} uses mutual recursion.  The function
\code{count\_down} is similar to the function with the same name in
example \emph{t37}.  However, we do not count down to $0$ but to a
variable $y$ that is passed as an argument and we call the function
\code{count\_up} afterwards.  The function \code{count\_up} is dual to
\code{count\_down}.  Here, we count up $y$ by $2$ and recursively call
\code{count\_down}.  For the function $\_main$, which calls
\code{count\_down(y,z)}, the analyzer computes the tight bound $1.67 +
1.33 |[z,y]|$.

@@

$XXX$
\\[.7\baselineskip]
      {\bf Greatest Common Divisor}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
 XXX
   \end{lstlisting}

$XXX$
\\[.7\baselineskip]
      {\bf Quick Sort}
    \end{center}
  \end{minipage}
